---
title: "Introduction to the landmarkR package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the landmarkR package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tidyverse)
library(landmarkR)
```

# Setup

In this vignette, we use the dataset epileptic to perform landmarking analysis of time-to-event data with time-varying covariates. Here is the structure of the dataset.

```{r}
data("epileptic")
str(epileptic)
```

The dataset contains the following variables:

-   id: a unique patient identifier

-   time: time when time-varying covariate "dose" was recorded

-   with.time: time when the first of event or censoring happened

-   with.status: indicates whether event (1) or censoring (0) occurered

-   dose: a time-varying covariate

-   treat, age, gender, learn.dis: static (baseline) covariates

The first step is to split the dataset into two, one containing static covariates, event time and indicator of

event/censoring, and another one containing dynamic covariates.

```{r}
# DF with Static covariates
df_static <- epileptic |>
  select(id, with.time, with.status, treat, age, gender, learn.dis) |>
  rename(patient_id = id, event.time = with.time, event.status = with.status) |>
  unique()
df_static$age <- (df_static$age - mean(df_static$age)) / sd(df_static$age)

head(df_static)
```

```{r}
# DF with Dynamic covariates
df_dynamic <- epileptic |>
  select(id, time, dose) |>
  mutate(covariate = "dose") |>
  rename(patient_id = id, times = time, measurements = dose, covariates = covariate)

head(df_dynamic)
```

# Create Landmarking object

Next step is to create an object of class Landmarking, using the helper function of the same name.

```{r}
landmarking_object <- Landmarking(
  data_static = df_static,
  data_dynamic = df_dynamic,
  event_indicator = "event.status",
  dynamic_covariates = "dose",
  ids = "patient_id",
  event_time = "event.time",
  times = "times",
  measurements = "measurements",
  dynamic_covariate_names = "covariates"
)
```

Arguments to the helper function are the following:

-   data_static and data_dynamic: the two datasets that were just created.

-   event_indicator: name of the column that indicates the censoring indicator in the static dataset.

-   dynamic_covariates: array column names in the dynamic dataset indicating time-varying covariates.

-   ids: name of the column that identifies patients in both datasets.

-   event_time: name of the column that identifies time of event/censoring in the static dataset.

# Baseline survival analysis (without time-varying covariates)

First, we perform a survival without time-varying covariates. We can use this as a baseline to evaluate the performance of a subsequent landmark analysis with such covariates. First step is to establish the landmark times, and to work out the risk sets at each of those landmark times.

```{r}
landmarking_object <- landmarking_object |>
  compute_risk_sets(landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25))

landmarking_object
```

Now we call the function fit_survival to fit the survival model.

```{r}
landmarking_object <- landmarking_object |>
  fit_survival(
    landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25),
    horizons = seq(from = 2 * 365.25, to = 6 * 365.25, by = 365.25),
    method = "coxph",
    dynamic_covariates = c()
  )

landmarking_object@survival_fits
```

# Landmarking analysis with lme4 + coxph

Now we use the package lme4 to fit a linear mixed model of the time-varying covariate, dose. This first step is followed by fitting a Cox PH submodel using the longitudinal predictions as covariates.

```{r}
landmarking_object <- Landmarking(
  data_static = df_static,
  data_dynamic = df_dynamic,
  event_indicator = "event.status",
  dynamic_covariates = "dose",
  ids = "patient_id",
  event_time = "event.time",
  times = "times",
  measurements = "measurements",
  dynamic_covariate_names = "covariates"
)

landmarking_object <- landmarking_object |>
  compute_risk_sets(
    landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25)
  )
```

```{r}
landmarking_object <- landmarking_object |>
  fit_longitudinal(
    landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25),
    method = "lme4",
    formula = measurements ~ treat + age + gender + learn.dis + (times | patient_id)
  ) |>
  predict_longitudinal(
    landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25),
    method = "lme4",
    allow.new.levels = TRUE
  ) |>
  fit_survival(
    landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25),
    horizons = seq(from = 2 * 365.25, to = 6 * 365.25, by = 365.25),
    method = "coxph",
    dynamic_covariates = c("dose")
  )

landmarking_object@survival_fits
```

# Landmarking analysis with lcmm + coxph

Now we use the package lcmm to fit a latent class mixed model of the time-varying covariate, dose. This first step is followed by fitting a Cox PH submodel using the longitudinal predictions as covariates.

```{r}
landmarking_object <- Landmarking(
  data_static = df_static,
  data_dynamic = df_dynamic,
  event_indicator = "event.status",
  dynamic_covariates = "dose",
  ids = "patient_id",
  event_time = "event.time",
  times = "times",
  measurements = "measurements",
  dynamic_covariate_names = "covariates"
)
```

```{r}
landmarking_object <- landmarking_object |>
  compute_risk_sets(seq(from = 365.25, to = 5 * 365.25, by = 365.25)) |>
  fit_longitudinal(
    landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25),
    method = "lcmm",
    formula = measurements ~ treat + age + gender + learn.dis,
    mixture = ~ treat + age + gender + learn.dis,
    subject = "patient_id",
    ng = 2
  ) |>
  predict_longitudinal(
    landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25),
    method = "lcmm",
    subject = "patient_id",
    avg = FALSE
  ) |>
  fit_survival(
    landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25),
    horizons = seq(from = 2 * 365.25, to = 6 * 365.25, by = 365.25),
    method = "coxph",
    dynamic_covariates = c("dose")
  )
```
