---
title: "Introduction to the landmarkR package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the landmarkR package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tidyverse)
library(landmarkR)
```

# Setup

In this vignette, we use the dataset epileptic to perform landmarking analysis of time-to-event data with time-varying covariates. Here is the structure of the dataset.

```{r}
data("epileptic")
str(epileptic)
```

The dataset contains the following variables:

-   id: a unique patient identifier

-   time: time when time-varying covariate "dose" was recorded

-   with.time: time when the first of event or censoring happened

-   with.status: indicates whether event (1) or censoring (0) occurred

-   dose: a time-varying covariate

-   treat, age, gender, learn.dis: static (baseline) covariates

The first step is to split the dataset into two, one containing static 
covariates, event time and indicator of event/censoring, and another one 
containing dynamic covariates. To that end, we use the function split_wide_df.
That function returns a named list with the following elements:
- Under the name df_static, a dataframe containing static  covariates, event 
time and indicator of event/censoring.
- Under the name df_dynamic, a named list of dataframes, associating dynamic 
covariates to dataframes in long format containing longitudinal measurement of 
the relevant dynamic covariate.


```{r}
# DF with Static covariates
epileptic_dfs <- split_wide_df(
  epileptic,
  ids = "id",
  times = "time",
  static = c("with.time", "with.status", "treat", "age", "gender", "learn.dis"),
  dynamic = c("dose"),
  measurement_name = "value"
)
static <- epileptic_dfs$df_static
head(static)
```

```{r}
# DF with Dynamic covariates
dynamic <- epileptic_dfs$df_dynamic
head(dynamic[["dose"]])
```

# Create Landmarking object

Next step is to create an object of class Landmarking, using the helper function of the same name.

```{r}
landmarking_object <- Landmarking(
  data_static = static,
  data_dynamic = dynamic,
  event_indicator = "with.status",
  ids = "id",
  event_time = "with.time",
  times = "time",
  measurements = "value"
)
```

Arguments to the helper function are the following:

-   data_static and data_dynamic: the two datasets that were just created.

-   event_indicator: name of the column that indicates the censoring indicator in the static dataset.

-   dynamic_covariates: array column names in the dynamic dataset indicating time-varying covariates.

-   ids: name of the column that identifies patients in both datasets.

-   event_time: name of the column that identifies time of event/censoring in the static dataset.

# Baseline survival analysis (without time-varying covariates)

First, we perform a survival without time-varying covariates. We can use this as a baseline to evaluate the performance of a subsequent landmark analysis with such covariates. First step is to establish the landmark times, and to work out the risk sets at each of those landmark times.

```{r}
landmarking_object <- landmarking_object |>
  compute_risk_sets(landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25))

landmarking_object
```

Now we call the function fit_survival to fit the survival model.

```{r}
landmarking_object <- landmarking_object |>
  fit_survival(
    landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25),
    windows = seq(from = 1 * 365.25, to = 2 * 365.25, by = 365.25),
    method = "coxph",
    dynamic_covariates = c()
  ) |> predict_survival(
    landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25),
    windows = seq(from = 1 * 365.25, to = 2 * 365.25, by = 365.25),
    method = "coxph",
    type = "survival"
  )

performance_metrics(
  landmarking_object,
  landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25),
  windows = seq(from = 1 * 365.25, to = 2 * 365.25, by = 365.25)
)

```

# Landmarking analysis with lme4 + coxph

Now we use the package lme4 to fit a linear mixed model of the time-varying covariate, dose. This first step is followed by fitting a Cox PH submodel using the longitudinal predictions as covariates.

```{r}
landmarking_object <- Landmarking(
  data_static = static,
  data_dynamic = dynamic,
  event_indicator = "with.status",
  ids = "id",
  event_time = "with.time",
  times = "time",
  measurements = "value"
)

landmarking_object <- landmarking_object |>
  compute_risk_sets(
    landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25)
  )
```

```{r}
landmarking_object <- landmarking_object |>
  fit_longitudinal(
    landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25),
    method = "lme4",
    formula = value ~ treat + age + gender + learn.dis + (time | id)
  ) |>
  predict_longitudinal(
    landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25),
    method = "lme4",
    allow.new.levels = TRUE
  ) |>
  fit_survival(
    landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25),
    windows = seq(from = 1 * 365.25, to = 2 * 365.25, by = 365.25),
    method = "coxph",
    dynamic_covariates = c("dose")
  ) |>
  predict_survival(
    landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25),
    windows = seq(from = 1 * 365.25, to = 2 * 365.25, by = 365.25),
    method = "coxph", 
    type = "survival"
  )

performance_metrics(
  landmarking_object,
  landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25),
  windows = seq(from = 1 * 365.25, to = 2 * 365.25, by = 365.25)
)
```

# Landmarking analysis with lcmm + coxph

Now we use the package lcmm to fit a latent class mixed model of the time-varying covariate, dose. This first step is followed by fitting a Cox PH submodel using the longitudinal predictions as covariates.

```{r}
landmarking_object <- Landmarking(
  data_static = static,
  data_dynamic = dynamic,
  event_indicator = "with.status",
  ids = "id",
  event_time = "with.time",
  times = "time",
  measurements = "value"
)
```

```{r}
landmarking_object <- landmarking_object |>
  compute_risk_sets(seq(from = 365.25, to = 5 * 365.25, by = 365.25)) |>
  fit_longitudinal(
    landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25),
    method = "lcmm",
    formula = value ~ treat + age + gender + learn.dis,
    mixture = ~ treat + age + gender + learn.dis,
    subject = "id",
    ng = 2
  ) |>
  predict_longitudinal(
    landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25),
    method = "lcmm",
    subject = "id",
    avg = FALSE
  ) |>
  fit_survival(
    landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25),
    windows = seq(from = 1 * 365.25, to = 2 * 365.25, by = 365.25),
    method = "coxph",
    dynamic_covariates = c("dose")
  ) |> 
  predict_survival(
    landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25),
    windows = seq(from = 1 * 365.25, to = 2 * 365.25, by = 365.25),
    method = "coxph", 
    type = "survival"
  )

```
